---
title: "IPD Meta-analysis"
author: "YJ"
date: "2025-10-25"
output:
  html_document:
    df_print: paged
---

### Data Loading
```{r packages_loading}
# Packages loading
library(lme4)         # (one-stage model)
library(metafor)      # (two-stage model)
library(broom.mixed) 
library(performance) 
library(influence.ME)
library(ggplot2)
library(dplyr)
library(purrr)
library(lmerTest)    # use it for Satterthwaite CI
```

```{r data_loading}
# Data Loading
data_factor_2trt <- readRDS( "D://Dissertation Dataset//data_factor_2trt.rds") %>%
  mutate(
    trial = as.factor(trial),
    sex_bin = as.factor(sex_bin),
    smoking_bin = as.factor(smoking_bin),
    treatment01   = ifelse(treatment == "Intensive", 1L, 0L)
  )
head(data_factor_2trt)
```

### Initial one-stage model
```{r initial_model}
# With random intercept and slope
initial_mod <- lmer(pasdas_po ~ treatment01 + pasdas_bsl + age + BMI + 
               sex_bin + disease_duration + smoking_bin + 
               (1 + treatment01 | trial), 
               data = data_factor_2trt, REML = TRUE)

summary(initial_mod)
```

### We find initial model singular so we consider the sanity check model, which is a simpler model.
```{r sanity_model}
# sanity check model (random intercept only)
sanity_mod <- lmer(pasdas_po ~ treatment01 + pasdas_bsl + age + BMI + 
               sex_bin + disease_duration + smoking_bin + 
               (1 | trial), 
               data = data_factor_2trt, REML = TRUE)

summary(sanity_mod)
```
### The sanity model achieves convergence and is non-singular, so then move to diagnostics check
```{r diagnostics_check}
# Diagnostic plots and checks for model assumptions
diagnostics <- function(model) {
  par(mfrow = c(2, 2))
  
  # Residuals vs fitted values - check for homoscedasticity
  plot(fitted(model), residuals(model), 
       main = "Residuals vs Fitted Values",
       xlab = "Fitted Values", ylab = "Residuals")
  abline(h = 0, col = "red")
  
  # Q-Q plot for normality - check residual normality
  qqnorm(residuals(model), main = "Q-Q Plot of Residuals")
  qqline(residuals(model), col = "red")
  
  # Residual histogram - check distribution shape
  hist(residuals(model), main = "Distribution of Residuals",
       xlab = "Residuals")
  
  # Random effects diagnostics - check normality of random intercepts
  re_int <- ranef(model)$trial
  qqnorm(re_int[,"(Intercept)"], 
         main = "Q-Q Plot of Random Intercepts")
  qqline(re_int[,"(Intercept)"], col = "red")
  }


# Run diagnostics on final model
diagnostics(sanity_mod)
```

### Influential Analysis
```{r influence}
# Check for influential trials using Cook's distance and DFBETAS
influence_analysis <- function(model) {
  par(mfrow = c(1, 1))
  
  tryCatch({
    # Calculate Cook's distance
    infl <- influence(model, group = "trial")
    cook_values <- cooks.distance(infl)
    
    # Ensure cook_values has names
    if(is.null(names(cook_values))) {
      if(!is.null(rownames(infl$alt.fixed))) {
        names(cook_values) <- rownames(infl$alt.fixed)
      } else {
        names(cook_values) <- paste0("Trial_", 1:length(cook_values))
      }
    }
    
    print("Cook's Distance by Trial:")
    print(cook_values)
    
    # Calculate Cook's distance threshold
    cook_threshold <- 4 / length(cook_values)
    
    # Plot Cook's distance with trial names
    plot(1:length(cook_values), cook_values, type = "h", 
         main = "Cook's Distance by Trial",
         ylab = "Cook's Distance", xlab = "",
         lwd = 2, xaxt = "n")  
    
    # Add trial names as x-axis labels
    axis(1, at = 1:length(cook_values), labels = names(cook_values), las = 2)
    
    # Add threshold line and legend
    abline(h = cook_threshold, col = "red", lty = 2, lwd = 2)
    legend("topright", 
           legend = paste("Threshold =", round(cook_threshold, 3)),
           col = "red", lty = 2, lwd = 2,
           bg = "white")
    
    # Mark trials exceeding threshold
    influential_cook <- which(cook_values > cook_threshold)
    if (length(influential_cook) > 0) {
      points(influential_cook, cook_values[influential_cook], 
             col = "red", pch = 16, cex = 1.5)
      text(influential_cook, cook_values[influential_cook],
           labels = names(cook_values)[influential_cook],
           pos = 3, col = "red", font = 2)
    }
    
    
    # DFBETAS for treatment effect - check stability of treatment estimate
    dfb <- dfbetas(infl)
    if ("treatment01" %in% colnames(dfb)) {
      dfb_values <- dfb[, "treatment01"]
      
      # Ensure dfb_values has names
      if(is.null(names(dfb_values))) {
        names(dfb_values) <- names(cook_values)  # Use same names as cook_values
      }
      
      # Calculate DFBETAS threshold
      dfb_threshold <- 2 / sqrt(nobs(model))
      
      # Plot DFBETAS with trial names
      plot(1:length(dfb_values), dfb_values, type = "h",
           main = "DFBETAS for Average Treatment Effects",
           ylab = "DFBETAS", xlab = "",
           lwd = 2, xaxt = "n")  
      
      # Add trial names as x-axis labels
      axis(1, at = 1:length(dfb_values), labels = names(dfb_values), las = 2)
      
      # Add threshold lines and legend
      abline(h = c(-1, 1) * dfb_threshold, col = "red", lty = 2, lwd = 2)
      abline(h = 0, col = "gray", lty = 1)
      legend("topright", 
             legend = paste("Threshold = ±", round(dfb_threshold, 3)),
             col = "red", lty = 2, lwd = 2,
             bg = "white")
      
      # Mark trials exceeding threshold
      influential_dfb <- which(abs(dfb_values) > dfb_threshold)
      if (length(influential_dfb) > 0) {
        points(influential_dfb, dfb_values[influential_dfb], 
               col = "red", pch = 16, cex = 1.5)
      }
    }
    
  }, error = function(e) {
    cat("Influence analysis failed:", e$message, "\n")
  })
}

# Run influence analysis on sanity model
influence_analysis(sanity_mod)
```


### Sensitivity Analysis
```{r sensitivity}
# Sensitivity analysis: assess robustness by systematically removing each trial
sensitivity <- function(model, data) {
  
  trials <- unique(data$trial)
  sensitivity_results <- data.frame()
  
  # Fit model while removing each trial one at a time
  for (trial_to_remove in trials) {
    subset_data <- data %>% filter(trial != trial_to_remove)
    
    tryCatch({
      sens_model <- update(model, data = subset_data)
      
      fixed_effects <- fixef(sens_model)
      if ("treatment01" %in% names(fixed_effects)) {
        result <- data.frame(
          Removed_Trial = trial_to_remove,
          Treatment_Effect = fixed_effects["treatment01"],
          Tau = as.data.frame(VarCorr(sens_model))[1, "sdcor"]
        )
        sensitivity_results <- rbind(sensitivity_results, result)
      }
    }, error = function(e) {
      cat("Model failed when removing trial", trial_to_remove, "\n")
    })
  }
  
  print(sensitivity_results)
  
  # Visualize sensitivity analysis results
  if (nrow(sensitivity_results) > 0) {
    p <- ggplot(sensitivity_results, aes(x = Removed_Trial, y = Treatment_Effect)) +
      geom_point(size = 3) +
      geom_hline(yintercept = fixef(model)["treatment01"], 
                 linetype = "dashed", color = "red") +
      labs(title = "Sensitivity Analysis: Treatment Effect When Removing Each Trial",
           y = "Treatment Effect", x = "Removed Trial") +
      theme_minimal()
    print(p)
  }
  
  return(sensitivity_results)
}

# Execute sensitivity analysis on final model
sensitivity_results <- sensitivity(sanity_mod, data_factor_2trt)
```

### View the summary statistics of sanity model and compare the result of Wald CI and Satterthwaite CI
```{r summary}
# Get all coefficients from summary table
summary_model <- summary(sanity_mod)
coefficients <- summary_model$coefficients

ate_estimate <- coefficients["treatment01", "Estimate"]
ate_se <- coefficients["treatment01", "Std. Error"]
ate_df <- coefficients["treatment01", "df"]

# Wald 95% CI
wald_ci <- confint(sanity_mod, method = "Wald", level = 0.95)
ate_wald_lower <- ate_estimate - 1.96 * ate_se
ate_wald_upper <- ate_estimate + 1.96 * ate_se

cat("\n=== Wald 95% CI ===\n")
cat(sprintf("ATE: %.3f (95%% CI: [%.3f, %.3f])\n", 
            ate_estimate, ate_wald_lower, ate_wald_upper))

# Satterthwaite 95% CI
ci_satt <- data.frame(
  Estimate = coefficients[, "Estimate"],
  Lower_CI = coefficients[, "Estimate"] - qt(0.975, df = 3) * coefficients[, "Std. Error"],
  Upper_CI = coefficients[, "Estimate"] + qt(0.975, df = 3) * coefficients[, "Std. Error"],
  df = coefficients[, "df"],
  `Pr(>|t|)` = coefficients[, "Pr(>|t|)"]
)
print(ci_satt)
```


### Two-stage model setup; Making forest plot and funnel plot
```{r two_stage}
# Two-stage ipdma
# Stage 1: fit model by trial to get ATE respectively
two_stage_results <- data_factor_2trt %>%
  group_by(trial) %>%
  do({
    fit <- lm(
      pasdas_po ~ treatment01 + pasdas_bsl + age + BMI + sex_bin + 
        disease_duration + smoking_bin,
      data = .
    )
    coef_summary <- summary(fit)$coefficients
    data.frame(
      trial = unique(.$trial),
      ATE = coef_summary["treatment01", "Estimate"],
      SE = coef_summary["treatment01", "Std. Error"]
    )
  })

print(two_stage_results)

# Stage 2: pool results using random-effects meta-analysis
meta_res <- rma(yi = ATE, sei = SE, data = two_stage_results, method = "REML",test="knha")
summary(meta_res)

# Weight of each trial
w <- weights(meta_res)
w_perc <- 100 * w / sum(w)          
w_lab  <- sprintf("%.2f", w_perc)           

# Adjust the size of the squares shown in the plot according to weight
psize_vector <- 0.8 + w / 30

# Add p-value, I^2, and tau^2 estimate info
mlabfun <- function(text, x) {
   list(bquote(paste(.(text),
       " (",
      .(fmtp2(x$QEp)), ", ",
      I^2, " = ", .(fmtx(x$I2, digits=1)), "%, ",
      tau^2, " = ", .(fmtx(x$tau2, digits=2)), ")")))}

pred <- predict(meta_res)

# Forest plot
forest(meta_res, 
       header = c("Study", "Mean Difference [95% CI]"),
       psize = psize_vector,
       efac = 2,
       col = "blue",
       slab = two_stage_results$trial,
       xlab = "Mean Difference in PASDAS Score (Intensive vs. Standard)",
       mlab = mlabfun("Two-stage Summary",meta_res),
       ilab = w_lab, 
       ilab.lab = "two-stage weight%",
       ilab.xpos = -2.5,         
       ylim = c(-3, 6.5),
       cex = 0.9)               

# Add one-stage result to the plot
ate_ci_satt <- ci_satt["treatment01", c("Estimate", "Lower_CI", "Upper_CI")]
addpoly(x = ate_ci_satt$Estimate,
        ci.lb = ate_ci_satt$Lower_CI,
        ci.ub = ate_ci_satt$Upper_CI,
        col = "green",
        rows = -2,
        cex = 0.9,
        mlab = "One-stage Summary") 

# Add a red vertical line across two-stage diamond
abline(v = meta_res$b,      
       col = "red",         
       lty = "dashed",      
       lwd = 1)             

# Make a funnel plot
funnel(meta_res, main="Standard Error",  xlab = "Treatment effect (MD)", legend=list(show="cis"))
```

### Save the result of one stage model for later use
```{r save_one_stage}
# Save the data since we will use it again in another rmd
one_stage_vals <- list(
  ATE      = ate_ci_satt$Estimate,
  CI_lower = ate_ci_satt$Lower_CI,
  CI_upper = ate_ci_satt$Upper_CI
)

dir.create("outputs", showWarnings = FALSE)
saveRDS(one_stage_vals, file = "outputs/one_stage_ate.rds")
```

















